The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2018-10-15 article.tex
Correctly converts to PDF

\usepackage[english]{babel}

@@

\usepackage{pgf}
\usepackage{hyperref}
\usepackage{verbatim}

@@

\item In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the common per-species speciation and extinction rates present also in the standard BD model. 
Additionally, MBD relies on two additional parameters. Parameter $\nu$ is the rate at which an environmental change is triggered.
When such event is triggered, all species present in the phylogeny at that moment
have a probability $q$ to speciate at that time, which is 
independent on $\lambda$.

@@

\item We start simulating $N_{S} = 1000$ \richel{I will measure the number of trees we'll be able to simulate within a short enough time, when the experiment is set up} MBD trees. From each MBD tree, a DNA sequence alignment is simulated. For each sequence alignment we then perform a Bayesian analysis to recover a posterior distribution of trees, each composed of $N_{P}$ phylogenies. Such analysis is performed using 
the 'pirouette' package (\cite{pirouette}) to call the BEAST2 tool 
suite from R. We let the Bayesian analysis assume a BD prior, to investigate
the error this inference makes due to this.

\item For each tree 
generated under the MBD model we aim to generate a "twin" tree under the BD model in order to perform a fair comparison. 
We want trees from the two models to contain the same amount of information, 
i.e. the same (expected) number of DNA mutations and 
the same number of taxa at the present.
To obtain these twin trees, 
\richel{
  I suggest to first start describing how to create the twin tree,
  calculation of the speciation rates, then the twin alignment.
}
\gio{@richel: both assumptions seem to make sense to me. 
  If possible I would run both and see which one performs better. For now
  let's stick to option 1 in the manuscript.
}
\richel{
  Sorry for the confusion: the previous comment was about the order
  of describing out methods in this manuscript. As of the methods,  
  I really meant to create twin trees with 
  both equal number of taxa and equal expected number of mutations.
}

@@

The expected number of mutations $m$ of a phylogeny 
\richel{
  I think 'expected number of mutations' would be more correct.
  Do you agree?
}
\gio{
  I don't agree. It is expected if you use an expected number 
  of species over time. If you have the correct $n(t)$ in principle you 
  should be able to get the exact number of mutations.
}
\richel{
  If you have the correct $n(t)$ you can indeed calculate exactly what
  the number of mutations will be, if and only if the mutation rate is
  zero. Except for this travial case, I am curious about how you
  calculate this. What is the precise number of mutations on a DNA
  sequence of 1000 base pairs over one time unit with an 
  expected average per-nucleotide
  mutation rate of once per time? I would say one expects on average 1000 mutations
  to occur. What would your claim be?
}

@@

\item So far we have simulated two datasets of trees under the two models: 
$\{T_{i}^{BD}\}_{i=1}^{N_{S}}$ and $\{T_{i}^{MBD}\}_{i=1}^{N_{S}}$.
We used them to generate a dataset of alignments for each model: $\{X^{BD}_{i}\}_{i=1}^{N_{S}}$ and $\{X^{MBD}_{i}\}_{i=1}^{N_{S}}$. From each dataset we produced a posterior distribution from a BD prior: 
$P_{i}(\theta | X^{BD}_{i}, BD)$ and $P_{i}(\theta | X^{MBD}_{i}, BD)$.
\gio{
  1) We might want to rename the models, e.g. BD = (0) and MBD = (1). These names with capital letters are too big and ugly;
}
\richel{
  I definitely get your point! 
  I would prefer something expressive over a number, perhaps a fancy letter, like
  for example $\mathbb{B}$ and $\mathbb{M}$ for a Birth-Death and Multiple-Birth Death. 
}
\gio{
2) Writing down things to tidy up manuscript and thoughts I just realized that 
  actually the "model" is the same for the two posteriors. 
  In fact with the word "model" Maturana et al. refer to the prior 
  used (see eq. 1, $\pi(\theta | M)$). In such case are we sure we can actually 
  use the Bayes factor for model selection?
}
\richel{
  Correct! The assumed model of the posterior is a Birth-Death model. To do
  model selection, we could assume a different nucleotide substitution 
  model (GTR instead of JC69) and/or a different clock model (relaxed clock,
  instead of a strict clock). I hope this clears things up. I will add
  'generative model' for the model (BD or MBD) that created a phylogeny
  and 'inferential model' for the model used in generating a posterior.
}

@@

$nLTT_{i,j} = \int_{0}^{T} \abs{N_{i,j}(t) - N_{T_{i}}} dt$

@@

$$
\delta nLTT = \int_{0}^{1} \abs{nLTT_1(t) - nLTT_2{t}} dt$
$$

>>>>

\usepackage{amssymb}
\usepackage[english]{babel}

@@

\usepackage{pgf}
\usepackage{hyperref}
\usepackage{verbatim}
\usepackage{comment}

@@

\item In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the common per-species speciation and extinction rates present also in the standard BD model. 
Additionally, MBD relies on two additional parameters. Parameter $\nu$ is the rate at which an environmental change is triggered.
When such event is triggered, all species present in the phylogeny at that moment
have a probability $q$ to speciate at that time, which is 
independent on $\lambda$. Polytomies are not allowed in such process as each species can speciate only once at the time.

@@

\item We start simulating $N_{S} = 1000$ \richel{I will measure the number of trees we'll be able to simulate within a short enough time, when the experiment is set up} MBD trees. From each MBD tree, a DNA sequence alignment is simulated. For each sequence alignment we then perform a Bayesian analysis to recover a posterior distribution of trees, each composed of $N_{P}$ phylogenies. Such analysis is performed using 
the 'pirouette' package (\cite{pirouette}) to call the BEAST2 tool 
suite from R. We let the Bayesian analysis assume a BD prior in both cases, to investigate
the extent of the error we make under this assumption.

\item For each tree 
generated under the MBD model we aim to generate a "twin" tree under the BD model. With the word "twin" we denote a tree generated starting from the respective MBD tree, in order to perform a fair comparison with it. This operation has to be done because we want to compare two trees that are generated by different processes. To do so we infer the parameters $\lambda_{BD}$ and $\mu_{BD}$ from the MBD maximizing the likelihood under a BD model. To perform this operation we use the function "\texttt{bd\_ML}" from the package "\texttt{DDD}" (\cite{etienne2012diversity}).

\item We then exploit such parameters to generate a BD tree using the function "\texttt{tess.sim.taxa.age}" from the package "\texttt{TESS}" (\cite{Hoehna2013}). We simulate the tree in such a way the new tree has the same number of tips and the same crown age as the MBD tree. We furthermore require that the BD tree conserve the topology of the MBD tree

\gio{see the razzo pipeline file: I am not sure what is the best way to do so}.

\begin{comment}
We want trees from the two models to contain the same amount of information, 
i.e. the same (expected) number of DNA mutations and 
the same number of taxa at the present.
To obtain these twin trees, 
\end{comment}
\richel{
  I suggest to first start describing how to create the twin tree,
  calculation of the speciation rates, then the twin alignment.
}
\gio{@richel: both assumptions seem to make sense to me. 
  If possible I would run both and see which one performs better. For now
  let's stick to option 1 in the manuscript.
}
\richel{
  Sorry for the confusion: the previous comment was about the order
  of describing out methods in this manuscript. As of the methods,  
  I really meant to create twin trees with 
  both equal number of taxa and equal expected number of mutations.
}
\gio{The procedure is explained in the razzo pipeline markdown file. See there. CLOSED}

@@

The expected number of mutations $m$ of a phylogeny 
\richel{
  I think 'expected number of mutations' would be more correct.
  Do you agree?
}
\gio{
  I don't agree. It is expected if you use an expected number 
  of species over time. If you have the correct $n(t)$ in principle you 
  should be able to get the exact number of mutations.
}
\richel{
  If you have the correct $n(t)$ you can indeed calculate exactly what
  the number of mutations will be, if and only if the mutation rate is
  zero. Except for this trivial case, I am curious about how you
  calculate this. What is the precise number of mutations on a DNA
  sequence of 1000 base pairs over one time unit with an 
  expected average per-nucleotide
  mutation rate of once per time? I would say one expects on average 1000 mutations
  to occur. What would your claim be?
}
\gio{Now I agree. CLOSED}

@@

\item So far we have simulated two datasets of trees under the two models: 
$\{T_{i}^{BD}\}_{i=1}^{N_{S}}$ and $\{T_{i}^{MBD}\}_{i=1}^{N_{S}}$.
We used them to generate a dataset of alignments for each model: $\{X^{BD}_{i}\}_{i=1}^{N_{S}}$ and $\{X^{MBD}_{i}\}_{i=1}^{N_{S}}$. From each dataset we produced a posterior distribution from a BD prior: 
$P_{i}(\theta | X^{BD}_{i}, BD)$ and $P_{i}(\theta | X^{MBD}_{i}, BD)$.
\gio{
  1) We might want to rename the models, e.g. BD = (0) and MBD = (1). These names with capital letters are too big and ugly;
}
\richel{
  I definitely get your point! 
  I would prefer something expressive over a number, perhaps a fancy letter, like
  for example $\mathbb{B}$ and $\mathbb{M}$ 
  for a Birth-Death and Multiple-Birth Death. 
}
\gio{
2) Writing down things to tidy up manuscript and thoughts I just realized that 
  actually the "model" is the same for the two posteriors. 
  In fact with the word "model" Maturana et al. refer to the prior 
  used (see eq. 1, $\pi(\theta | M)$). In such case are we sure we can actually 
  use the Bayes factor for model selection?
}
\richel{
  Correct! The assumed model of the posterior is a Birth-Death model. To do
  model selection, we could assume a different nucleotide substitution 
  model (GTR instead of JC69) and/or a different clock model (relaxed clock,
  instead of a strict clock). I hope this clears things up. I will add
  'generative model' for the model (BD or MBD) that created a phylogeny
  and 'inferential model' for the model used in generating a posterior.
}
}

@@

$nLTT_{i,j} = \int_{0}^{T} | N_{i,j}(t) - N_{T_{i}} | dt$

@@

$$
\Delta nLTT = \int_{0}^{1} | nLTT_1(t) - nLTT_2{t} | dt
$$
